---
title: "Draft: Data Documentation"
---

```{r}
library(readr)
library(purrr)

monthly_files <- list.files(pattern = "gom_A5B7_t\\d{2}mn01\\.csv")

monthly_dfs <- map(monthly_files, ~ read_csv(.x, show_col_types = FALSE))

monthly_dfs <- map(monthly_files, ~ read_csv(
  .x,
  col_names = c("Latitude", "Longitude", "Temp_C"),
  skip = 2,
  col_types = cols_only(
    Latitude = col_double(),
    Longitude = col_double(),
    Temp_C = col_double()
  )
))

dauphin_clean <- function(df, month_num) {
  df %>%
    filter(
      between(Latitude, 29.5, 30.5),
      between(Longitude, -88.5, -87.5)
    ) %>%
    mutate(
      Temp_F = Temp_C * 9/5 + 32,
      Above_85 = Temp_F > 85,
      Month = month_num
    )
}
```

The statistical mean is the average of all unflagged interpolated values at each standard depth level for each variable in each 1° square which contains at least one measurement for the given oceanographic variable. Uses longitude and latitude to tell location.

Source: https://www.ncei.noaa.gov/access/gulf-of-mexico-climate/bin/gomregcl.pl

Dauphin Island 30, 88

```{r}
df2 <- readxl::read_xlsx("2024_ESD.xlsx")
```

NERC collects, maintains, and annually publishes the Electricity Supply and Demand (ES&D) database, which includes 10-year projections for the North American bulk power system (BPS). Data are presented for each NERC Region (and subregions where applicable) for years 1990-2010. For years 2011 to present, data are presented for each NERC Assessment Area. Some information in the ES&D has been aggregated to protect sensitive information.